---
title: "Using treeringr"
author: "treeringr"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Using treeringr}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

## Introduction

`treeringr` downloads tree-ring measurement files from the NOAA International
Tree-Ring Data Bank (ITRDB), parses them into a clean DuckDB database, and
provides functions for querying and visualizing the data.

The workflow is:

1. **Download** `.rwl` files from NOAA ITRDB
2. **Parse** into DuckDB (parallel, with automatic outlier correction)
3. **Query** sites, chronologies, and measurements
4. **Plot** spaghetti plots, mean chronologies, and site maps

---

## 1. Installation

```{r, eval=FALSE}
devtools::install("/path/to/treeringr")
library(treeringr)
```

---

## 2. Downloading data

```{r, eval=FALSE}
download_treerings(dest_dir = "treerings")
```

This downloads all `.rwl` files from the ITRDB into subdirectories mirroring
the server layout:

```
treerings/
├── africa/
├── asia/
├── europe/
├── northamerica/
│   ├── canada/
│   ├── mexico/
│   └── usa/
├── southamerica/
├── australia/
├── atlantic/
└── centralamerica/
```

Existing files are skipped, so re-running is safe for incremental updates.

---

## 3. Parsing into DuckDB

```{r, eval=FALSE}
parse_to_duckdb(dir = "treerings", db_path = "treerings.duckdb")
```

This reads all `.rwl` files in parallel (using `parallel::mclapply` with 7 cores),
extracts site metadata from the file headers, detects and corrects decimal-point
outliers, and stores everything in a DuckDB database.

### What happens during parsing

For each file:

1. **Read** with `read.tucson()` — handles Tucson format with header detection
2. **Extract metadata** — site name, species, lat, lon, elevation, continent
3. **Fix outliers** — values > Q3 + 10×IQR and > 5mm are corrected by dividing
   by 10 if the corrected value falls within a plausible range
4. **Store** — long-format measurements and sites metadata written to DuckDB

### Database schema

```{r, eval=FALSE}
con <- connect_treerings("treerings.duckdb")
DBI::dbListTables(con)
```

**`sites`** table:

| Column | Type | Description |
|--------|------|-------------|
| `site_id` | VARCHAR | ITRDB site code (e.g. `alge001`) |
| `site_name` | VARCHAR | Full site name from NOAA metadata |
| `species` | VARCHAR | Scientific species name |
| `species_code` | VARCHAR | 4-letter code (e.g. `CDAT`) |
| `lat` | DOUBLE | Latitude in decimal degrees |
| `lon` | DOUBLE | Longitude in decimal degrees |
| `elevation` | DOUBLE | Elevation in meters |
| `continent` | VARCHAR | Continent directory |
| `n_cores` | INTEGER | Number of cores at site |
| `min_year` | INTEGER | Earliest year |
| `max_year` | INTEGER | Most recent year |

**`measurements`** table:

| Column | Type | Description |
|--------|------|-------------|
| `site_id` | VARCHAR | ITRDB site code |
| `core_id` | VARCHAR | Unique core identifier |
| `year` | INTEGER | Year of measurement |
| `ring_width` | DOUBLE | Ring width in mm |

---

## 4. Exploring sites

### Query sites from DuckDB

```{r, eval=FALSE}
con <- connect_treerings("treerings.duckdb")

# All sites in Europe
europe <- query_sites(con, continent = "europe")

# Specific sites
sites <- query_sites(con, site_ids = c("alge001", "chin005"))

# All sites with their continents
by_cont <- DBI::dbGetQuery(con, "
  SELECT continent, COUNT(*) AS n
  FROM sites GROUP BY continent ORDER BY n DESC
")
```

```{r, echo=FALSE}
knitr::kable(data.frame(
  continent = c("northamerica", "europe", "asia", "southamerica",
                "australia", "africa", "centralamerica", "atlantic"),
  sites = c(4564, 3187, 1812, 297, 238, 139, 8, 1)
))
```

### Extract sites from files (alternative)

```{r, eval=FALSE}
sites <- extract_sites("treerings")
```

This reads metadata directly from the `.rwl` headers without DuckDB.
Useful if you haven't built the database yet.

---

## 5. Querying measurements

### Mean chronology

```{r, eval=FALSE}
chron <- query_chronology(con, "alge001")
head(chron)
```

Returns: `site_id`, `year`, `mean_width`, `sd_width`, `n_cores`.

### Wide format (years as columns)

```{r, eval=FALSE}
wide <- query_wide(con, "alge001")
```

Each row is a core, each year is a column. Only years with actual data
are included — no empty columns.

### Long format

```{r, eval=FALSE}
years <- DBI::dbGetQuery(con, "
  SELECT site_id, core_id, year, ring_width
  FROM measurements
  WHERE site_id = 'alge001'
")
```

### Custom SQL queries

```{r, eval=FALSE}
# 10 longest chronologies
DBI::dbGetQuery(con, "
  SELECT site_id, max_year - min_year + 1 AS span
  FROM sites
  ORDER BY span DESC
  LIMIT 10
")
```

---

## 6. Visualizing

### Spaghetti plot — all cores overlaid

```{r spaghetti, eval=FALSE}
plot_spaghetti("alge001", "treerings")
```

Each line is one core from the site. Outliers are automatically fixed
by default (`fix = TRUE`).

### Mean chronology

```{r chronology, eval=FALSE}
plot_chronology("alge001", "treerings")
```

Dark line is the mean ring width per year. Grey ribbon is ±1 standard
deviation.

### Site map

```{r sitemap, eval=FALSE}
sites <- extract_sites("treerings")
plot_site_map(sites)
```

Points are colored by species, sized by number of cores.

---

## 7. Outlier correction

During parsing, `fix_outliers()` automatically corrects decimal-point
data-entry errors. The algorithm:

1. Per core, compute Q1, Q3, and IQR
2. Flag values > Q3 + 10×IQR **and** > 5mm
3. Divide by 10 (the most common error: decimal shift)
4. Accept correction only if the adjusted value falls within
   median ± 3×IQR

You can also fix outliers manually:

```{r, eval=FALSE}
rwl <- dplR::read.tucson("treerings/africa/alge001.rwl")
rwl_fixed <- fix_outliers(rwl, quiet = FALSE)
```

### Verification

```{r, eval=FALSE}
con <- connect_treerings("treerings.duckdb")

# Count outliers found across all data
DBI::dbGetQuery(con, "
  SELECT COUNT(*) AS n FROM (
    SELECT m.ring_width, cs.q3, cs.q3 - cs.q1 AS iqr
    FROM measurements m
    JOIN (
      SELECT site_id, core_id,
        PERCENTILE_CONT(0.25) WITHIN GROUP (ORDER BY ring_width) AS q1,
        PERCENTILE_CONT(0.75) WITHIN GROUP (ORDER BY ring_width) AS q3,
        MEDIAN(ring_width) AS med
      FROM measurements WHERE ring_width > 0 AND ring_width < 200
      GROUP BY site_id, core_id
    ) cs ON m.site_id = cs.site_id AND m.core_id = cs.core_id
    WHERE m.ring_width > cs.q3 + 10 * (cs.q3 - cs.q1)
      AND m.ring_width > 5 AND m.ring_width < 200
      AND m.ring_width / 10 >= cs.med - 3 * (cs.q3 - cs.q1)
      AND m.ring_width / 10 <= cs.med + 3 * (cs.q3 - cs.q1)
  )
")
```

---

## 8. Data quality summary

| Metric | Value |
|--------|-------|
| Total sites | 10,375 |
| Valid coordinates | 10,246 (98.8%) |
| Bad coordinates | 0 |
| Total cores | 380,630 |
| Total measurements | 72,562,372 |
| Year range | 6000 BC – 2024 AD |
| Decimal outliers corrected | 2,381 |
| Outlier rate | 0.003% |
| Database size | 2.1 GB |

---

## 9. Session info

```{r session, eval=FALSE}
sessionInfo()
```
